Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines

### 2025-07-03 docs/paper.Rmd
write theoretical background (recent refinements - intrahousehold bargaining)s)

>>>>

A further refinement of the quantity–quality (Q–Q) model considers the question of *who* within the household makes fertility and child investment decisions. The original Beckerian framework adopts a unitary model of the family, and assumes a single utility function and complete agreement between spouses over optimal fertility \( n \) and child quality \( q \). In practice, however, empirical evidence reveals significant heterogeneity in preferences between household members along gender lines. According to \citet{oppenheim1987impact, thomas1990intra}, for instance, men often desire more children than women and may differ in their willingness to invest in each child’s education or health. These discrepancies have motivated game-theoretic models of intra-household bargaining, in which fertility and investment outcomes reflect the relative influence of each parent’s preferences.

In such models, the mother is typically assumed to have a stronger preference for child quality—such as health and schooling—while the father may favor either more children or alternative uses of household resources. The resolution of these conflicting preferences depends on bargaining power, which can be shaped by income contributions, legal rights, cultural norms, or access to external resources. When the mother’s bargaining power increases, theoretical models predict a shift toward lower fertility and higher per-child investment, holding other factors constant. This prediction is consistent with empirical findings: \citet{iyigun2007endogenous, doepke2019bargaining} show that greater female empowerment—via education or labor force participation—correlates with reduced fertility and increased investment in child human capital.

Mathematically, the household’s first-order condition for fertility in a bargaining model can be written as:

\[
\alpha \frac{\partial U_{\text{wife}}}{\partial n} + (1 - \alpha) \frac{\partial U_{\text{husband}}}{\partial n} = \lambda(p_n + p_q q),
\]

with a corresponding condition for \( q \). Here, \( \alpha \in [0,1] \) represents the wife’s bargaining weight. When \( \alpha \) increases, the composite marginal utility of additional children typically decreases—especially if the wife prefers fewer children—leading to lower equilibrium fertility and a shift along the Q–Q frontier toward higher quality.

Recent work also explores dynamic bargaining, in which spouses negotiate sequential decisions over time, potentially leading to strategic behavior (e.g., one partner may accelerate or delay subsequent births). Although these models introduce complexity, their core implication for the Q–Q framework is clear: household power dynamics fundamentally shape the trade-off between child quantity and quality. In societies where women have limited decision-making autonomy—due to lack of access to contraception, or social norms—fertility tends to remain high and per-child investment low, which stalls demographic transition. Conversely, when women gain bargaining power—through legal reforms, labor market participation, or targeted transfers—the household often reallocates resources toward fewer but higher-quality children.

This theoretical insight is corroborated by policy experiments. For instance, cash transfer programs directed to mothers consistently lead to greater spending on children’s health, education, and nutrition compared to equivalent transfers given to fathers \citep{duflo2003grandmothers, thomas1990intra} . Such outcomes support the hypothesis that mothers place higher weight on child quality, and that who controls the purse strings matters deeply. In sum, introducing intra-household bargaining into the Q–Q model enriches its explanatory scope: it highlights how family structure and power asymmetries—not just income levels or prices—generate variation in fertility and human capital outcomes across and within societies.

\subsection{Health and Nutrition in the Q–Q Model}

### 2025-07-03 docs/paper.tex
write theoretical background (recent refinements - intrahousehold bargaining)s)

\date{\today}
\pubMonth{06}
\pubYear{2025}
\pubVolume{}
\pubIssue{}
\JEL{J13, I15, O15}
\Keywords{fertility shocks, child nutrition, quantity-quality
trade-off, urban poverty, Philippines}

@@

>>>>

\date{\today}
\pubMonth{07}
\pubYear{2025}
\pubVolume{}
\pubIssue{}
\JEL{J13, I15, O15}
\Keywords{fertility shocks, child nutrition, quantity-quality
trade-off, urban poverty, Philippines}

@@

A further refinement of the quantity--quality (Q--Q) model considers the
question of \emph{who} within the household makes fertility and child
investment decisions. The original Beckerian framework adopts a unitary
model of the family, and assumes a single utility function and complete
agreement between spouses over optimal fertility \(n\) and child quality
\(q\). In practice, however, empirical evidence reveals significant
heterogeneity in preferences between household members along gender
lines. According to \citet{oppenheim1987impact, thomas1990intra}, for
instance, men often desire more children than women and may differ in
their willingness to invest in each child's education or health. These
discrepancies have motivated game-theoretic models of intra-household
bargaining, in which fertility and investment outcomes reflect the
relative influence of each parent's preferences.

In such models, the mother is typically assumed to have a stronger
preference for child quality---such as health and schooling---while the
father may favor either more children or alternative uses of household
resources. The resolution of these conflicting preferences depends on
bargaining power, which can be shaped by income contributions, legal
rights, cultural norms, or access to external resources. When the
mother's bargaining power increases, theoretical models predict a shift
toward lower fertility and higher per-child investment, holding other
factors constant. This prediction is consistent with empirical findings:
\citet{iyigun2007endogenous, doepke2019bargaining} show that greater
female empowerment---via education or labor force
participation---correlates with reduced fertility and increased
investment in child human capital.

Mathematically, the household's first-order condition for fertility in a
bargaining model can be written as:

\[
\alpha \frac{\partial U_{\text{wife}}}{\partial n} + (1 - \alpha) \frac{\partial U_{\text{husband}}}{\partial n} = \lambda(p_n + p_q q),
\]

with a corresponding condition for \(q\). Here, \(\alpha \in [0,1]\)
represents the wife's bargaining weight. When \(\alpha\) increases, the
composite marginal utility of additional children typically
decreases---especially if the wife prefers fewer children---leading to
lower equilibrium fertility and a shift along the Q--Q frontier toward
higher quality.

Recent work also explores dynamic bargaining, in which spouses negotiate
sequential decisions over time, potentially leading to strategic
behavior (e.g., one partner may accelerate or delay subsequent births).
Although these models introduce complexity, their core implication for
the Q--Q framework is clear: household power dynamics fundamentally
shape the trade-off between child quantity and quality. In societies
where women have limited decision-making autonomy---due to lack of
access to contraception, or social norms---fertility tends to remain
high and per-child investment low, which stalls demographic transition.
Conversely, when women gain bargaining power---through legal reforms,
labor market participation, or targeted transfers---the household often
reallocates resources toward fewer but higher-quality children.

This theoretical insight is corroborated by policy experiments. For
instance, cash transfer programs directed to mothers consistently lead
to greater spending on children's health, education, and nutrition
compared to equivalent transfers given to fathers
\citep{duflo2003grandmothers, thomas1990intra} . Such outcomes support
the hypothesis that mothers place higher weight on child quality, and
that who controls the purse strings matters deeply. In sum, introducing
intra-household bargaining into the Q--Q model enriches its explanatory
scope: it highlights how family structure and power asymmetries---not
just income levels or prices---generate variation in fertility and human
capital outcomes across and within societies.
